Dynamics of Laterally Propagating Flames in X-ray Bursts. II. Realistic Burning \& Rotation

### 2021-02-25 xrb2/paper.tex
address more referee comments

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We continue to investigate two-dimensional laterally propagating flames in type I X-ray bursts using fully compressible hydrodynamics simulations.  In the current study we relax previous approximations where we artificially boosted the flames. We now use more physically realistic reaction rates, thermal conductivities, and rotation rates, exploring the effects of neutron star rotation rate and thermal structure on the flame.  We find that at lower rotation rates the flame becomes harder to ignite, whereas at higher rotation rates the nuclear burning is enhanced by increased confinement from the Coriolis force and the flame propagates steadily. At higher crustal temperatures, the flame moves more quickly and accelerates as it propagates through the atmosphere. If the temperature is too high, instead of a flame propagating across the surface the entire atmosphere burns steadily. All of the software used for these simulations is freely available.

@@

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. We saw in \citet{flame_wave1} that 
increasing the neutron star rotation rate reduces the horizontal lengthscale of the flame. An 
estimate of this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius 
may be thought of as the scale over which the balance between the Coriolis force and horizontal 
pressure gradient becomes important, and is approximately given by
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is the gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use $\dot{e}_\mathrm{nuc}$ measured at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ to discern the horizontal extent of the flame at different rotation rates. Taking the edge at greatest radius of the bright teal/green region where the most significant energy generation is occurring as the leading edge of the flame in each plot, we see that the horizontal extent of the $1000~\mathrm{Hz}$ flame ($\tt{F1000}$) appears to be 
reduced compared to the lower rotation $500~\mathrm{Hz}$ run ($\tt{F500}$). \added{As the natural aspect ratio makes it hard to see the flame structure, we show the energy generation rate for the $\tt{F1000}$ simulation with the vertical extent stretched in Figure~\ref{fig:flame_stretch}.}  From Equation 
\ref{eqn:rossby}, we can see that increasing the rotation rate from $500~\mathrm{Hz}$ to 
$1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and that the greater confinement from 
the Coriolis force should reduce the horizontal extent of the flame by a similar factor. However, 
the Rossby radius is only an approximate measure of this horizontal lengthscale, and in our 
simulations we see that this scaling does not work so well for all rotation rates. The simulations 
seem to follow the theoretical scaling described in Equation \ref{eqn:rossby} more closely at 
higher rotation rates ($1000~\mathrm{Hz}$ and higher), based on the results of \citet{flame_wave1}.

@@

\added{For the simple case of a laminar flame front driven by conduction, the propagation rate, 
$s_l$, of the flame is approximately
\begin{equation}
	\label{eqn:laminar_flame_speed}
	s_l \approx \left( \frac{\mathcal{D} \enucdot}{E} \right)^{\half}.
\end{equation}
Here, $\mathcal{D} = k_{th} / (\rho c_v)$ is the thermal diffusivity, $\enucdot$ is the specific
nuclear energy generation rate, and $E$ is the specific total energy \citep{timmeswoosley:1992}. 
\citet{cavecchi:2013} explored the effects of rotation on the flame propagation, and noted that the 
flame speed in their simulations scaled with the ratio $L_R/H_0$ of Rossby length to atmospheric 
scale height. Combining this scaling with Equation \ref{eqn:laminar_flame_speed} yields a revised 
estimate for the flame speed, $s$:
\begin{equation}
	s \approx \frac{L_R}{H_0} s_l \approx \frac{L_R}{H_0} \left( \frac{\mathcal{D} \enucdot}{E} 
	\right)^{\half}.
\end{equation}
Observing that $L_R$ is inversely proportional to the angular rotation frequency $\Omega$ and 
and neglecting variations in $H_0$ and $E$, we can then derive the following approximate scaling 
relation:
\begin{equation}
	\label{eqn:flame_speed_scaling}
	s \appropto \frac{(\mathcal{D} \enucdot)^{\half}}{\Omega}
\end{equation}
We were able to test the validity of this scaling relation in \citet{flame_wave1}, where we studied 
the effects of varying the thermal conductivity/diffusivity, energy generation rate, and rotation 
rate. These results supplemented the findings of \citet{cavecchi:2013}. Despite this formulation of 
the flame speed neglecting some relevant physics (e.g.\ turbulence), the relation proved
consistent with our simulation results.}

@@

\acknowledgments \castro\ is open-source and freely available at
\url{http://github.com/AMReX-Astro/Castro}.  The problem setup used
here is available in the git repo as {\tt
  flame\_wave}.  \added{We thank the anonymous referee for the helpful suggestions.}
  The work at Stony Brook was supported by DOE/Office
of Nuclear Physics grant DE-FG02-87ER40317.  This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Advanced
Scientific Computing Research and Office of Nuclear Physics, Scientific
Discovery through Advanced Computing (SciDAC) program under Award
Number DE-SC0017955.  This research was supported by the Exascale Computing Project (17-SC-20-SC), a collaborative effort of the U.S. Department of Energy Office of Science and the National Nuclear Security Administration.
This material is also based upon work supported by the U.S. Department
of Energy, Office of Science, Office of Advanced Scientific Computing Research, Department of
Energy Computational Science Graduate Fellowship under Award Number DE-SC0021110. This work was supported in part by the U.S. Department of Energy, Office of Science, Office of Workforce Development for Teachers and Scientists (WDTS) under the Science Undergraduate Laboratory Internship (SULI) program. MZ acknowledges support from the Simons Foundation. 
This research used resources of the National Energy
Research Scientific Computing Center, a DOE Office of Science User
Facility supported by the Office of Science of the U.~S.\ Department
of Energy under Contract No.\ DE-AC02-05CH11231.  This research used
resources of the Oak Ridge Leadership Computing Facility at the Oak
Ridge National Laboratory, which is supported by the Office of Science
of the U.S. Department of Energy under Contract No. DE-AC05-00OR22725,
awarded through the DOE INCITE program.  We thank NVIDIA Corporation
for the donation of a Titan X and Titan V GPU through their academic
grant program.  This research has made use of NASA's Astrophysics Data
System Bibliographic Services.

>>>>

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
We continue to investigate two-dimensional laterally propagating flames in type I X-ray 
bursts using fully compressible hydrodynamics simulations.  In the current study we relax 
previous approximations where we artificially boosted the flames. We now use more physically 
realistic reaction rates, thermal conductivities, and rotation rates, exploring the effects 
of neutron star rotation rate and thermal structure on the flame.  We find that at lower 
rotation rates the flame becomes harder to ignite, whereas at higher rotation rates the 
nuclear burning is enhanced by increased confinement from the Coriolis force and the 
flame propagates steadily. At higher crustal temperatures, the flame moves more quickly 
and accelerates as it propagates through the atmosphere. If the temperature is too high, 
instead of a flame propagating across the surface the entire atmosphere burns 
\replaced{steadily}{uniformly}. \added{Our findings could have implications for the 
relationship between observed burst rise times and neutron star rotation and accretion rates.} 
All of the software used for these simulations is freely available.

@@

We run three models ({\tt F250}, {\tt F500}, and {\tt F1000}) with the same initial model in terms 
of temperature but differing rotation rates. We saw in \citet{flame_wave1} that 
increasing the neutron star rotation rate reduces the horizontal lengthscale of the flame. An 
estimate of this lengthscale is given by the Rossby radius of deformation, $L_R$. The Rossby radius 
may be thought of as the scale over which the balance between the Coriolis force and horizontal 
pressure gradient becomes important, and is approximately given by
\begin{equation}
	\label{eqn:rossby}
	L_R \approx \frac{\sqrt{g H_0}}{\Omega},
\end{equation}
where $g$ is the gravitational acceleration, $H_0$ is the atmospheric scale height, and $\Omega$ is the 
neutron star rotation rate. \added{If we take $g \sim 3.6 \times 10^{12}~\mathrm{cm}~\mathrm{s}^{-2}$ and $H_0\sim 10^3~\mathrm{cm}$, then for the $500~\mathrm{Hz}$ system we find the Rossby length to approximately be $L_R \sim 1.2 \times 10^5~\mathrm{cm}$. This estimate is likely to be an overestimate of the true value of the Rossby length in our simulations, as we saw that the flame is confined on a much smaller length scale.} In Figure \ref{fig:time_series_enuc_500} and Figure 
\ref{fig:time_series_enuc_1000}, we use $\dot{e}_\mathrm{nuc}$ measured at $50~\mathrm{ms}$ and $100~\mathrm{ms}$ to discern the horizontal extent of the flame at different rotation rates. Taking the edge at greatest radius of the bright teal/green region where the most significant energy generation is occurring as the leading edge of the flame in each plot, we see that the horizontal extent of the $1000~\mathrm{Hz}$ flame ($\tt{F1000}$) appears to be 
reduced compared to the lower rotation $500~\mathrm{Hz}$ run ($\tt{F500}$). \added{As the natural aspect ratio makes it hard to see the flame structure, we show the energy generation rate for the $\tt{F1000}$ simulation with the vertical extent stretched in Figure~\ref{fig:flame_stretch}.}  From Equation 
\ref{eqn:rossby}, we can see that increasing the rotation rate from $500~\mathrm{Hz}$ to 
$1000~\mathrm{Hz}$ should decrease $L_R$ by a factor of two, and that the greater confinement from 
the Coriolis force should reduce the horizontal extent of the flame by a similar factor. However, 
the Rossby radius is only an approximate measure of this horizontal lengthscale, and in our 
simulations we see that this scaling does not work so well for all rotation rates. The simulations 
seem to follow the theoretical scaling described in Equation \ref{eqn:rossby} more closely at 
higher rotation rates ($1000~\mathrm{Hz}$ and higher), based on the results of \citet{flame_wave1}.

@@

\added{For the simple case of a laminar flame front driven by conduction, the propagation rate, 
$s_l$, of the flame is approximately
\begin{equation}
	\label{eqn:laminar_flame_speed}
	s_l \approx \left( \frac{\mathcal{D} \enucdot}{E} \right)^{\half}.
\end{equation}
Here, $\mathcal{D} = k_{th} / (\rho c_v)$ is the thermal diffusivity, $\enucdot$ is the specific
nuclear energy generation rate, and $E$ is the specific total energy \citep{timmeswoosley:1992}. 
\citet{cavecchi:2013} explored the effects of rotation on the flame's propagation, and noted that the 
flame speed in their simulations scaled with the ratio $L_R/H_0$ of Rossby length to atmospheric 
scale height. Combining this scaling with Equation \ref{eqn:laminar_flame_speed} yields a revised 
estimate for the flame speed, $s$:
\begin{equation}
	s \approx \frac{L_R}{H_0} s_l \approx \frac{L_R}{H_0} \left( \frac{\mathcal{D} \enucdot}{E} 
	\right)^{\half}.
\end{equation}
Observing that $L_R$ is inversely proportional to the angular rotation frequency $\Omega$ and 
and neglecting variations in $H_0$ and $E$, we can then derive the following approximate scaling 
relation:
\begin{equation}
	\label{eqn:flame_speed_scaling}
	s \appropto \frac{(\mathcal{D} \enucdot)^{\half}}{\Omega}.
\end{equation}
We were able to test the validity of this scaling relation in \citet{flame_wave1}, where we studied 
the effects of varying the thermal conductivity/diffusivity, energy generation rate, and rotation 
rate. These results supported the findings of \citet{cavecchi:2013}. Despite this formulation of 
the flame speed neglecting some relevant physics (e.g.\ turbulence), the relation proved
consistent with our simulation results.}

@@

\acknowledgments \castro\ is open-source and freely available at
\url{http://github.com/AMReX-Astro/Castro}.  The problem setup used
here is available in the git repo as {\tt
  flame\_wave}.  \added{We thank the anonymous referee for their helpful suggestions.}
  The work at Stony Brook was supported by DOE/Office
of Nuclear Physics grant DE-FG02-87ER40317.  This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Advanced
Scientific Computing Research and Office of Nuclear Physics, Scientific
Discovery through Advanced Computing (SciDAC) program under Award
Number DE-SC0017955.  This research was supported by the Exascale Computing Project (17-SC-20-SC), a collaborative effort of the U.S. Department of Energy Office of Science and the National Nuclear Security Administration.
This material is also based upon work supported by the U.S. Department
of Energy, Office of Science, Office of Advanced Scientific Computing Research, Department of
Energy Computational Science Graduate Fellowship under Award Number DE-SC0021110. This work was supported in part by the U.S. Department of Energy, Office of Science, Office of Workforce Development for Teachers and Scientists (WDTS) under the Science Undergraduate Laboratory Internship (SULI) program. MZ acknowledges support from the Simons Foundation. 
This research used resources of the National Energy
Research Scientific Computing Center, a DOE Office of Science User
Facility supported by the Office of Science of the U.~S.\ Department
of Energy under Contract No.\ DE-AC02-05CH11231.  This research used
resources of the Oak Ridge Leadership Computing Facility at the Oak
Ridge National Laboratory, which is supported by the Office of Science
of the U.S. Department of Energy under Contract No. DE-AC05-00OR22725,
awarded through the DOE INCITE program.  We thank NVIDIA Corporation
for the donation of a Titan X and Titan V GPU through their academic
grant program.  This research has made use of NASA's Astrophysics Data
System Bibliographic Services.
